Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-04-05 WSSSPE1_report/WSSSPE1_summary_report.tex
improve the flow of text and cleanup

In this section, we discuss some of the software projects as case-studies to
better understand the points discussed during the workshop and described in the
previous sections. We dive deeper to find how are they affected by the
sustainability issues in practice. Additionally, We discuss various issues
including career paths of original developers and investigate if the original
developers are still around?

We classify the software projects discussed in the workshop in two broad
categories. First, the utility software, comprising of software that enable
and/or facilitate the development of other tools and techniques to carry out
scientific work. This includes the software developed to efficiently utilize
new research infrastructures. Second, scientific software, comprising software
that was developed with an aim to solve a specific scientific problem.

% Why are we classifying software like above?
This classification is motivated by our argument that the software projects
wildly vary in factors such as scope, purpose and usage. The development and
management of software is significantly influenced by these factors.
Consequently, the sustainability challenges faced by software differ and must
be treated separately. For instance, the challenges faced by a gateway software
development project such as \textit{CIPRES} or \textit{VisIT} are distinct to a
niche software for ab initio modeling and simulation such as \textit{VASP} or
\textit{Quantum Espresso}.

\subsection{Utility Software}

We classify software developed with a potentially wider audience and general
purpose usage in mind as Utility software. In the context of research, a
utility software could be defined as general purpose system applicable to one
or many generic tasks and/or enabling other software to run by providing a
suitable environment. Utility software typically do not address a fundamental
research challenge in a given scientific domain. Examples are collaborative
development frameworks such as \emph{GitHub} and \emph{Bitbucket}, distributed
and workflow and generic computing frameworks such as \emph{Galaxy},
\emph{nanoHUB}, \emph{SimGrid}, \emph{Swift}, \emph{Globus} and \emph{VisTrails}, and
visualization frameworks such as \emph{VisIT}.

% What are the sustainability challenges? What are the oppurtunities?
Development is often a high risk/reward undertaking and explores uncharted
territories and is largely dominated by (re)usability factors. Owing to
relatively large number of features, the development and prototype process is
also lengthy which poses a significant survival risk.
Ferenbaugh~\cite{Ferenbaugh_WSSSPE} further discusses the risks and challenges
associated with software projects dealing with new architectures.

On the other hand, owing to its generic nature, utility software presents
oppurtunities to be usable by a larger community. This makes it more visible
which in turn leads to a broader and deeper participation. For
instance, promoting collaborations across the breadth (e.g. different
departments) and depth (e.g. stakeholders within a department) of community,
one of the key ingredient of a sustainable process. Successful projects reap
high rewards and have longer usage span. Development process becomes
user-driven and self-sustaining.

One such example is the \emph{Galaxy} project. It follows agile software
development practices and implements standard practices such as test-driven
development and socialized bug managing practices via \emph{trello}. Galaxy
\emph{histories} and \emph{toolshed} offers easy community sharing of data and
tools further promoting a collaborative environment. The project very closely
follows the guidelines described in~\cite{Carver_WSSSPE} and many
from~\cite{Prlic_WSSSPE}. Many utility software projects are often developed
aiming better utilizing a particular, new infrastructure and architecture,
e.g., \emph{MVAPICH}, \emph{VisIT}. Similarly, to leverage the power of
emerging architectures such as CPU accelerators new code and libraries are
required. The experience of one such effort as described
in~\cite{Ferenbaugh_WSSSPE} met with a limited success but nonetheless with
many invaluable lessons were learned about influential cultural and technical
aspects in sustainable software development practices.

A relatively new paradigm in utility software is the software delivered as
service over the web. With increasing popularity of cloud-based storage and
computational environments, many users are leaning towards tools used as
services. \emph{GitHub} and \emph{Bitbucket} can be arguably considered to be
such tools, catering to collaborative development. For scientific users
\emph{Globus}-based tools are a case of service-based utility software
discussed during the workshop. The data movement and sharing services offered
by Globus can be easily used over the web by collaborating researchers.
%
%
%\note{Dan: so is the only difference here that this category is aimed at new
%systems, where utility software is not?  If so, perhaps this should be
%incorporated into that subsection, so there would just be two subsections.
%Something in the framing of this subsection doesn't feel right to me.  FL: I
%respectfully disagree the the only reason to develop frameworks is to better
%utilize (new or old) infrastructure. Another really important reason is to
%modularize development, making it much easier for an individual developer/user
%to integrate their own code. Ketan: I think in this section I wanted to bring
%up a class of software projects which are undertaken with the main aim of
%tackling new systems and architectures in mind. New infrastructures such as
%GPU, Cloud, fast IB, federal science networks, etc. were on my mind as I see
%many submissions around them in the workshop.}
%
%

\subsection{Scientific software}

Scientific software consists primarily of special-purpose software that was
purpose-built for a target use-case scenario or fixed/frozen requirements in
mind.  Software projects pertaining to specific scientific domain often tend to
be in a niche and the user community tends to be small to medium. They are
mostly driven by the science and specific needs of a research group.

% What are the challenges?
Long-term sustainability of scientific research software is often a significant
challenge. Many submissions reported that software is practically considered a
`byproduct' of the actual research. Others contended that the software was not
the main funded part of their research.  Specific needs such as numerical
accuracy and algorithmic optimization are some of the paramount requirements of
most scientific software.  The resulting codebase is relatively small. A
smaller codebase and fixed requirements result in stability, ease of
installation, and configurability.  Many such projects mature and are treated
as libraries to be used into larger systems such as some of the utility
software discussed in the previous section.  While the software can stay stable
and require relatively low maintenance, the responsibility is often on the
shoulders of a very few developers who might not be specialists in software
development. Development tends to be linear and simplistic with a limited
scope to follow software best practices.

% TODO: What are the oppurtunities?

Some examples of such software discussed as part of the workshop are,
\emph{DUNE}, \emph{R/qtl}, \emph{Kitware}, \emph{PETSc}, \emph{MINRES-QLP},
\emph{FASTMath}.  Most of the above efforts are focused on one scientific
and/or applied mathematics domain. However, sometimes such projects grow beyond
the initial vision of developers. One such example is \emph{Kitware}, which
while being a software product specializing in the scientific process, has a
core focus of developing communities around software processes. An instance of
this process is the development of the \emph{CMake} build utility, which
started out as a building tool for \emph{ITK} but grew to become a generic
build utility for C++ projects. Similarly, \emph{PETSc} is growing towards
becoming a general purpose utility system usable for solving variety of
scientific problems.
% 
In order to investigate the much discussed issue of career paths for developers
of scientific software, we conducted a small survey of the use-cases discussed
above with respect to the current status of its core developers. We looked at
the workshop submissions and the original cited papers for the respective
software. Where available, we looked at the webpages and source code of the
earliest versions to find the original authors. We excluded relatively new
projects such as \emph{SciPy} and \emph{numFOCUS} from this survey. we find
that in a majority of scientific software more than one original developers are
still around, in some cases for more than 10 years, (eg. \emph{R/qtl}).
Whereas, in the case of utility software projects, it appears that at least one
original developer is still around often in a part-time role. In other cases,
it appears that the original developers have went on to assume a larger role,
again with the same project.

In conclusion, we find that there are distinctions in the characteristics and
the kinds of challenges faced between generic utility and specific scientific
domain software projects. In terms of career paths of original developers, it
appears that the career paths for utility software developers are broader
compared to the developers of scientific software.  However, it appears that in
the case of scientific software, core competencies of original developers tend
to be domain science who happen to be the coders of their software. As projects
as a whole, the small-scale, specific software have a lower risk and cost in
terms of sustaining as compared to utility software.  Often the sustainability
of such projects is achieved by the fact that the core developer or team
heavily utilizes the software for their science, e.g.,  \emph{R/qtl, PETSc}.
Furthermore, the development of scientific software requires more scientific
skills compared with those of utility software. This diminishes the need of
specialist programmers and practices in many cases where the bulk of
development is carried out by a domain scientist. With these inferences, we
believe that separate guidelines and/or principles for sustainability be
defined and practiced for each category of scientific software and they be
given separate treatment.

>>>>

In this section, we discuss some of the software projects as case studies to
better understand the points discussed during the workshop and described in the
previous sections. We dive deeper to find how are they affected by the
sustainability related issues in practice. Additionally, we discuss issues
surrounding people, especially the career paths of original developers and
investigate to answer the question: are the original developers are still
around?

We classify the software projects discussed in the workshop in two broad
categories. First, the \emph{utility} software, comprising of general purpose
software. Utility software is often used as enabler and/or facilitator for the
development of other tools and techniques to carry out scientific work. This
includes the software developed to efficiently utilize new research
infrastructures. Second, the \emph{scientific} software, comprising the
software that was originally developed with an aim to solve a specific
scientific problem.

% Why are we classifying software like above?
This classification is motivated by our argument that the two kinds of software
projects wildly vary in factors such as scope, purpose and usage. The
development and management of each kind is significantly different.
Consequently, the sustainability challenges faced by them differ and must be
treated separately. For instance, the challenges faced by a gateway software
development project such as \textit{CIPRES} or a visualization software product
such as \textit{VisIT} are distinct to a niche software for \textit{ab initio}
modeling and simulation such as \textit{VASP} or \textit{Quantum Espresso}.

\subsection{Utility Software}
Software developed with a potentially wider audience and general purpose usage
in mind is utility software. Utility software typically does not address
fundamental research problems for a given scientific domain. Examples are
collaborative development frameworks such as \emph{GitHub} and
\emph{Bitbucket}, distributed workflow and generic computing frameworks
such as \emph{Galaxy}, \emph{nanoHUB}, \emph{SimGrid}, \emph{Swift},
\emph{Globus} and \emph{VisTrails}, and visualization frameworks such as
\emph{VisIT}.

% What are the sustainability challenges? What are the oppurtunities?
Development is often a high risk/reward undertaking exploring uncharted
territories and is largely influenced by (re)usability factors. Owing to
relatively large number of features, the development and prototype process is
also lengthy which poses a significant survival risk. Challenges on a class of
utility software for new architectures is well discussed
in~\cite{Ferenbaugh_WSSSPE}.

On the other hand, utility software presents oppurtunities to be usable by a
larger community making its undertaking and development an attractive pursuit.
It is generally more visible in community which in turn leads to a broader and
deeper participation. For instance, promoting collaborations across the breadth
(e.g.  different departments) and depth (e.g. stakeholders within a department)
of community, one of the key ingredient of a sustainable process. Successful
utility projects reap high rewards and have a longer usage span. Development
process becomes user-driven and self-sustaining.

One such example is the \emph{Galaxy} project. It follows agile software
development practices and implements standards such as test-driven development
and socialized bug managing practices via \emph{trello}. Galaxy
\emph{histories} and \emph{toolshed} offers easy community sharing of data and
tools further promoting a collaborative environment. The project very closely
follows the guidelines described in~\cite{Carver_WSSSPE} and many
from~\cite{Prlic_WSSSPE}. Many utility software projects are often developed
aiming better utilizing a particular, new infrastructure and architecture,
e.g., \emph{MVAPICH}, \emph{VisIT}. Similarly, to leverage the power of
emerging architectures such as accelerators, new code and libraries are
required. The experience of one such effort as described
in~\cite{Ferenbaugh_WSSSPE} which met with a limited success but nonetheless
with many invaluable lessons were learned about influential cultural and
technical aspects in sustainable software development practices.

A relatively new paradigm in utility software is the software delivered as
service over the web. With increasing popularity of cloud-based storage and
computational environments, many users are leaning towards tools used as
services. \emph{GitHub} and \emph{Bitbucket} can be argued to be
such tools, catering to collaborative development. For scientific users
\emph{Globus}-based tools are a case of service-based utility software
discussed during the workshop. The data movement and sharing services offered
by Globus can be easily used over the web by collaborating researchers.
%
%
%\note{Dan: so is the only difference here that this category is aimed at new
%systems, where utility software is not?  If so, perhaps this should be
%incorporated into that subsection, so there would just be two subsections.
%Something in the framing of this subsection doesn't feel right to me.  FL: I
%respectfully disagree the the only reason to develop frameworks is to better
%utilize (new or old) infrastructure. Another really important reason is to
%modularize development, making it much easier for an individual developer/user
%to integrate their own code. Ketan: I think in this section I wanted to bring
%up a class of software projects which are undertaken with the main aim of
%tackling new systems and architectures in mind. New infrastructures such as
%GPU, Cloud, fast IB, federal science networks, etc. were on my mind as I see
%many submissions around them in the workshop.}
%
%

\subsection{Scientific software}
Scientific software consists primarily of special-purpose software that was
purpose-built for a target use-case scenario,  fixed/frozen requirements in
mind or solving a specific problem. Software projects pertaining to specific
scientific domain tend to be in a niche and the user community tends to be
small to medium. They are mostly driven by the science and specific needs of a
research group. Specific needs such as numerical accuracy and
algorithmic optimization are some of the paramount requirements of most
scientific software.

% What are the challenges?
Long-term sustainability of scientific software is often a significant
challenge and face radically different issues compared to utility software.
Many submissions reported that software is practically considered a `byproduct'
of the actual research. Others contended that the software was not the main
funded part of their research. A smaller codebase and fixed requirements result
in stability, ease of installation, and configurability.  Many such projects
mature and are treated as libraries to be used into larger systems such as some
of the utility software discussed in the previous section. While the software
can stay stable and require relatively low maintenance, the responsibility is
often on the shoulders of a very few developers who might not be specialists in
software development. Development tends to be linear and simplistic with a
limited scope to follow software best practices.

Some examples of such software discussed as part of the workshop are,
\emph{DUNE}, \emph{R/qtl}, \emph{Kitware}, \emph{PETSc}, \emph{MINRES-QLP},
\emph{FASTMath}.  Most of the above efforts are focused on one scientific
and/or applied mathematics domain. However, sometimes such projects grow beyond
the initial vision of developers. One such example is \emph{Kitware}, which
while being a software product specializing in the scientific process, has a
core focus of developing communities around software processes. An instance of
this process is the development of the \emph{CMake} build utility, which
started out as a building tool for \emph{ITK} but grew to become a generic
build utility for C++ projects. Similarly, \emph{PETSc} is growing towards
becoming a general purpose utility system usable for solving variety of
scientific problems.
%

Finally, it appears that the career options of developers from different kinds
of software projects differ significantly. In order to further investigate the
much discussed issue of career paths for developers of these software, we
conducted a small survey of the use-cases discussed above with respect to the
current status of its core developers. We looked at the workshop submissions
and the original cited papers for the respective software. Where available, we
looked at the webpages and source code of the earliest versions to find the
original developers. We included projects older than 5 years and excluded
relatively new projects such as \emph{SciPy} and \emph{numFOCUS} from this
survey. we find that in a majority of scientific software more than one
original developers are still around, in some cases for more than 10 years,
(eg. \emph{R/qtl}).  Whereas, in the case of utility software projects, it
appears that at least one original developer is still around, often in a
part-time or ``benefactor" role. In other cases, it appears that the original
developers have went on to assume a larger role, with the same project.

In conclusion, we find that there are distinctions in the characteristics and
the kinds of challenges faced between utility and scientific software projects.
In terms of career paths of original developers, it appears that the career
choices for utility software developers are broader compared to the developers
of scientific software. However, it appears that in the case of scientific
software, core competencies of original developers tend to be domain science
who happen to be the coders of their software. As projects as a whole, the
small-scale, specific software have a lower risk and cost in terms of
sustaining as compared to utility software. We also find that the utility
software tend to adhere more loyally to the many best practices discussed
during the workshop. Often the sustainability of scientific software projects
is achieved by the fact that the core developer or team heavily utilizes the
software for their science, e.g.,  \emph{R/qtl, PETSc}.  Furthermore, the
development of scientific software requires more scientific skills compared
with those of utility software. This diminishes the need of specialist
programmers and practices in many cases where the bulk of development is
carried out by a domain scientist. With these inferences, we believe that
separate guidelines and/or principles for sustainability be defined and
practiced for each category of scientific software and they be given separate
treatment.
